\begin{abstract}
This paper presents MARS, a Memory-Augmented Rust Shell: a terminal editor,
multiplexer, and LLM agent that ship as a single binary. The assistant failures that
matter most at a terminal are information gaps, not capability gaps: ``run the unit
tests'' is \code{cargo test} in this project and \code{npm test} in the next, and no
amount of model scale can recover which one the user means. MARS closes the gap by
living where the information lives. It captures the commands the user actually accepts,
retrieves them with plain BM25 over local files (no embeddings, no data leaving the
machine), and injects them into the prompt. Under a leakage-controlled protocol in
which the store holds only \emph{paraphrases} of earlier requests, this lifts a small
model (\code{claude-haiku-4-5}) from 38\% (9/24) to 96\% (23/24) on project-specific
commands, one short of the verbatim-lookup ceiling: generalization across a paraphrase
gap, not lookup of a planted answer. Pointed at the tool's
own documentation, the same mechanism raises accuracy on questions about the tool from
43\% to 76\% overall, with the decisive gain on plain-English reconfiguration (17\% to
93\%); the residual failures are informative, arising exactly where a strong parametric
prior overrides retrieved specifics. We describe the registry-grounded directive protocol, the corrective-memory loop,
the architecture that keeps the agent present across machines while the API key stays
home, and a self-instrumenting evaluation built from the agent's own debug log. MARS
is open source (MIT); a screencast accompanies this paper.
\end{abstract}
