\section{The Memory-Augmented Agent}
\label{sec:agent}

\input{sections/fig-architecture}

\subsection{Design considerations}
\label{sec:design}

\emph{The agent should be another input device, not a second system.} The recurring risk
in an agentic editor is that the agent becomes a subsystem bolted beside the editor, with
its own command language, state, and failure modes. We chose the opposite
design. Everything MARS can do is a variant of a single \code{Action} enum, and three
interfaces resolve a user's intent to it: a direct keychord (procedural recall), a fuzzy
command bar (recognition), and the agent (natural language). All three terminate in a
single dispatch point, so a capability added once is chord-bindable, searchable, and
agent-invokable. Figure~\ref{fig:arch} sketches the retrieval loop.

This choice trades expressive freedom for grounding. Because the agent can only name
actions that exist in the registry, and retrieves from flat local files rather than a
vector index, the surface a user must trust is small and easy to audit. However, MARS
cannot invent a capability at runtime, and lexical retrieval will miss paraphrases that
an embedding model would catch (\S\ref{sec:eval}). In the terminal domain we believe this trade is
strongly favorable.

\subsection{The agent as a validated input device}
\label{sec:directives}

\emph{An agent that acts on the user's workspace must be safer than a keypress, never a
bypass around the editor's gates.} Following the mode-error tradition in interface
design \citep{norman1988design}, we treat the agent as one input device among several,
behind the same confirmation gates as a keystroke. The implementation is a trailing-line
\emph{directive} protocol: the model receives a live catalog of every action,
description, and keybinding, and ends its reply with at most one directive
(Table~\ref{tab:directives}). We chose a plain-text protocol over provider-native
tool-calling APIs because it works identically across OpenAI-compatible endpoints, a
native Anthropic path, and local open models, and because it renders as human-readable
text that the user confirms before anything fires; the cost is that we give up the
stricter typing a schema provides.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Directive} & \textbf{Effect (gated by confirm)} \\
\midrule
\code{RUN: <Action>}   & fire a registry action \\
\code{TYPE: <cmd>}     & type a command into a shell pane \\
\code{OPEN: path:line} & jump to a file location \\
\code{NEED: <context>} & pull more context; re-ask once \\
\bottomrule
\end{tabular}
\caption{The directive protocol. A \code{RUN:} naming an action absent from the live
registry is rejected before it fires; a destructive action passes the same confirmation
gate a human would.}
\label{tab:directives}
\end{table}

The grounding comes from the registry check. A \code{RUN:} whose name does not resolve
against the live \code{Action} set is rejected before dispatch, so the model cannot
execute a hallucinated capability. The \code{NEED:} directive closes a different gap. Rather than
sending the full screen with every request, the model can ask for more context (a pane's
scrollback, another tab) on demand; MARS supplies it and re-asks exactly once, so no
retrieval loop can form.

\subsection{Natural-language shell translation}
\label{sec:translate}

The most frequent agentic request at a terminal is to turn English into the right
command. \emph{Translation should be assistive and reversible, never an automatic
execution.} The implementation lives on the command bar: the user presses
\code{Ctrl+Space} and types plain English (``find the biggest checkpoints here'');
MARS sends the request with on-screen context (working directory, recent output) to
the model cascade (\S\ref{sec:cascade}) and renders the proposed command at the cursor
to run, edit, or discard. The same translator also serves as the
command bar's fallback: plain text that matches no registry action is treated as a
natural-language command request (\S\ref{sec:system}). Translation by itself is a commodity
capability \citep{lin2018nl2bash,lin2017tellina}; what personalizes it is the memory we
describe next.

\subsection{Memory 1: the user's own commands}
\label{sec:m1}

\emph{The user's own accepted commands are the highest-value context available, and they
are local to the terminal by construction.} A generic model does not know the user's
commands. Asked to ``launch the training run,'' it produces a plausible
\code{python train.py} when this project in fact uses
\code{sbatch --gres=gpu:8 scripts/train.slurm}. The implementation is a corrective loop. Every time
the user accepts (or edits and runs) a translated command, MARS appends the
\code{(request, command)} pair to a local JSON-lines file, and it also reads recent shell
history from \code{\$HISTFILE}. On the next request, MARS ranks these entries against the
query with BM25 \citep{robertson2009bm25}, with the standard parameters $k_1{=}1.5$ and
$b{=}0.75$ and no index to build, and injects the top three pairs as few-shot exemplars,
together with up to three ranked lines of recent shell history.

The loop is corrective in the precise sense that a command the model got wrong once
becomes one it gets right thereafter, because the user's correction is what enters the
store; a per-invocation assistant never observes the correction. Nothing in this loop is
novel as information retrieval; the contribution is that in this domain retrieval this
simple is still decisive (\S\ref{sec:eval}), at a bounded cost of a few exemplars per
request.

\subsection{Memory 2: the tool's own documentation}
\label{sec:m2}

\emph{A tool's own capability surface is authoritative documentation and cannot drift
from the code.} An agent embedded in a tool should answer questions about that tool, and
change its settings, from ground truth rather than a guess. MARS's action registry,
keybindings, and tuning knobs each carry a human-written description, so documentation
is generated from the code itself. The same BM25 retriever runs over this second corpus:
documentation chunked into passages, plus the live registry and the described knobs. On a \code{NEED: docs} pull, MARS injects the top
passages, so ``how do I split the screen?'' is answered with the real keybinding and
``make the accent orange'' resolves to the actual knob and a confirmable edit. When
retrieval returns nothing, the agent reports that the capability does not exist rather
than inventing one.

\subsection{Why simple memory is enough}
\label{sec:whysimple}

It is worth stating why a method this simple suffices, because the instinct when
building a memory system is to reach for embeddings and a vector database. The terminal
is a constrained domain, and the constraint favors lexical retrieval in three ways.
Firstly, a user's command vocabulary is small, repetitive, and lexically stable: the same
flags, paths, and tool names recur verbatim, so a query and its useful memory tend to
share surface tokens, exactly the regime in which BM25 is strong and dense retrieval's
paraphrase advantage is marginal. Secondly, the corpora are small (hundreds of
entries), so there is no scaling pressure to justify an index.
Thirdly, the payoff is asymmetric: a single relevant exemplar of the user's own command
is worth more than semantic breadth, because it supplies the exact string the user wants
reproduced. We therefore treat embeddings as a cost to be deferred until the domain
broadens (\S\ref{sec:future}). The empirical support is the corrective-memory result
of \S\ref{sec:eval}, in which BM25 bridges a mean word-Jaccard gap of 0.36 with nothing
more than a prompt, a local file, and a standard ranking function.

\subsection{The model cascade}
\label{sec:cascade}

\emph{Model choice should be right-sized per task.} Most LLM traffic at a terminal is
cheap and can be served by small or free models, but a fixed model wastes quality
where it is not needed and hits rate limits where volume is high. We distinguish two
orthogonal reasons to change models. Rotating \emph{for limits} moves a task
horizontally, across models of the same tier, when a rate-limit error arrives; because
each provider meters usage on its own counter, rotating across $N$ families multiplies
the effective free ceiling by roughly $N$. Escalating \emph{for
quality} moves a task vertically to a stronger tier, and only on a detected failure.
Detection is free because the agent's output is structured: a
\code{RUN:} that fails the registry check is an unambiguous signal that a small model
erred, and it triggers one retry at a higher tier. The cascade costs a little dispatch
complexity; the benefit is headroom against the provider quota walls we later hit during
evaluation (\S\ref{sec:setup}).
