\section{System: The Substrate for Local Memory}
\label{sec:system}

Local memory is only useful if the agent is reliably present where the memory lives,
across disconnects and on machines the user does not own. The architecture exists to
guarantee that presence.

\paragraph{One binary, one core.}
MARS is a single Rust binary whose application core owns all state (buffers, panes,
terminals, the agent conversation, the memory stores) and never touches a TTY:
input arrives as source-neutral events, and rendering is a stateless projection onto a
\code{ratatui} backend. The same core runs unmodified as a standalone terminal, a
headless daemon, and a test backend, which lets the evaluation drive the real system
with no mocks (\S\ref{sec:eval}).

\paragraph{A persistent daemon.}
We implemented persistence as a process split rather than as a save-and-restore feature.
A server process runs the entire application headless and emits the editor's own ANSI
byte stream over a Unix socket; a thin client owns the real terminal and pumps bytes. We
chose to have the server render ANSI rather than define a structured wire protocol
because it reuses the entire rendering pipeline; the cost is a wire that carries
rendered frames rather than semantic events.
The benefit for memory is direct: because the core keeps running with no client attached,
the command store and the user's session survive a closed laptop, and an overnight run
can be watched and summarized for the user's return.

\paragraph{Classifier-free routing.}
The command bar routes on a \emph{sigil} rather than a learned classifier: the first
character of the input names the target namespace. \code{!} runs a shell command,
\code{?} asks the agent, \code{@} opens the file navigator, and plain text
fuzzy-searches the action registry. Plain text that matches nothing in the registry
falls back to the agent: the bar treats it as a natural-language command request and
hands it to the translator of \S\ref{sec:translate}, whose proposal the user confirms
before it runs. Routing remains deterministic under the fallback, triggered by a
registry miss rather than a classifier score, so there is no routing model to train,
evaluate, or explain.

\paragraph{The key never leaves home.}
\emph{A secret is a fact about the user, not the machine, and should live in one place.}
The implementation is a broker. A home-side daemon (\code{mars keyd})
holds the key; when the user runs \code{mars ssh}, MARS reverse-forwards the broker's
socket over the SSH connection, and the agent on the remote box makes its model calls by
proxying each request home, where the broker injects the credential and streams the
completion back. The key never lands on the remote machine, on disk, in the environment,
or in memory, so a compromised box has nothing to exfiltrate. The remaining exposure is
the open tunnel: a compromised box can use the credential while the tunnel is up, though
it cannot steal it, and access ends the moment the tunnel closes.

\paragraph{The log is the benchmark.}
Running with \code{--llm-debug} appends one JSON line per model call (task, model, real
token counts, latency, and the full prompt and reply) to a local stream, and records
whether each suggestion was accepted, edited, or rejected. The log is both a cost
profile the user can inspect and, as \S\ref{sec:eval} shows, the source of every prompt,
context, and token count the evaluation reports: the system instruments itself.
