\section{Evaluation}
\label{sec:eval}

\subsection{Setup}
\label{sec:setup}

We evaluate the two memory axes on frozen, a priori, author-written gold sets and a
single small model
under test: 112 command-translation items (88
general, 24 project-specific) and 82 self-knowledge question--answer items (52
knowledge, 30 reconfiguration); Axis A exercises the reflex memory
(\S\ref{sec:m1}), Axis B the reflective memory (\S\ref{sec:m2}). The model under test is \code{claude-haiku-4-5}: the thesis requires a \emph{small}
model, so that personalization has a gap to close, but not necessarily a free one. Two free tiers
(Qwen3-32B on Groq, Gemini Flash-Lite) exhausted their daily quotas mid-batch (the
wall the cascade of \S\ref{sec:cascade} routes around), so we ran the final numbers
on a small paid model; all reported runs had zero empty completions.

Correctness is scored by a stronger judge model, \code{claude-sonnet-5}, following
LLM-as-judge practice \citep{zheng2023judging}, so that the model under test does not
grade itself; judge and testee share a model family,
a limitation noted in the Limitations section.\footnote{As recorded in the log, the model identifiers are
\code{claude-haiku-4-5} (under test) and \code{claude-sonnet-5} (judge), both at
provider-default sampling (Claude 4.5-series models reject the temperature parameter);
the seed paraphrases come from a separate model pass, frozen before any run. Prompt
templates ship with the
source (\code{src/prompts/}) alongside the gold sets and result files.} During the experiment the reflex store contains the 24 seeded pairs and no
distractors; retrieval injects the top three (\S\ref{sec:m1}). Prompts,
contexts, and token counts come from the self-instrumenting log of \S\ref{sec:log}.

\subsection{Reflex memory: learning the user's commands}

\begin{table}[t]
\centering
\footnotesize
\begin{tabular}{@{}lc@{}}
\toprule
\textbf{Setting} & \textbf{Accuracy} \\
\midrule
General ($n{=}88$)        & 90\% (79/88) \\
Personalized ($n{=}24$)   & 42\% (10/24) \\
Overall ($n{=}112$)       & 79\% (89/112) \\
\midrule
\multicolumn{2}{@{}l}{\emph{Corrective memory ($n{=}24$, leakage-controlled)}} \\
\quad Cold (no memory)             & 38\% (9/24) \\
\quad Paraphrase memory            & \textbf{96\% (23/24)} \\
\quad Verbatim memory (ceiling)    & 100\% (24/24) \\
\bottomrule
\end{tabular}
\caption{Command translation (\code{claude-haiku-4-5}, judged by
\code{claude-sonnet-5}). The
corrective-memory block is leakage-controlled: the store holds only a \emph{paraphrase}
of each prior request and the model is tested on the original, so 96\% measures
generalization; verbatim memory is the lookup ceiling. Base and cold come from separate
runs that disagree on five items in both directions, netting to a one-item difference
(Appendix~\ref{app:examples}).}
\label{tab:axisA}
\end{table}

The base numbers in Table~\ref{tab:axisA} confirm the premise: the small model is a
competent general shell author at 90\% but falls to 42\% on project-specific
commands it cannot know. These are missing-knowledge failures,
chiefly \emph{ecosystem blindness} (Appendix~\ref{app:examples}), exactly what local
memory targets.

The corrective-memory experiment isolates that fix under a protocol that rules out mere
lookup. For each of the 24 tasks the store is seeded
not with the prior request verbatim but with an independently generated \emph{paraphrase}
(mean word-level Jaccard 0.36 to the test query), while the model is tested on the
\emph{original} request, so a warm win requires bridging the paraphrase gap, not copying a planted
answer. Under this setup,
cold accuracy is 38\% (9/24); paraphrase memory raises it to 96\% (23/24), with fifteen
items flipping to correct and one regressing from correct to wrong (the appendix
examines the flips and the regression). A verbatim-memory condition reaches the
lookup ceiling of 100\% (24/24). The
headline is the paraphrase number: simple BM25 over a local file recovers almost all of
the personalization gap by generalizing, not by lookup. Memory adds about 66 tokens per
call; the sign of that cost is model-dependent: on a reasoning model, the same
exemplars suppress chain-of-thought and net cost falls (Appendix~\ref{app:examples}).

\subsection{Reflective memory: learning the tool itself}

Retrieved documentation lifts overall accuracy on questions about MARS from 43\% to
76\% (Table~\ref{tab:axisB}), but the gain is unevenly distributed.
On ``how-do-I'' knowledge questions the gain is modest (58\% to 65\%): the no-memory
model already answers many correctly, and even with docs it sometimes hallucinates a
keybinding the docs state exactly (Appendix~\ref{app:examples}). The asymmetry has a
simple shape: a knob name is a token the retriever can hand to the model intact, while
a keybinding is a specific fact the model still overwrites with a more common default
(Appendix~\ref{app:examples}). The decisive gain is on \emph{reconfiguration}
(turning a request like ``autosave more often'' into a concrete knob edit), which
rises from 17\% to 93\%. Without the docs the model invents config filenames; with
the knob descriptions retrieved, it emits the exact \code{knob = value} in
\code{tuning.json}. One caveat, carried in the caption of Table~\ref{tab:axisB}:
part of the repair comes from the explain-don't-act instruction; the identifier-level
repair requires the retrieved text (Appendix~\ref{app:examples}).

\begin{table}[t]
\centering
\footnotesize
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Question type} & \textbf{No memory} & \textbf{+\,Docs memory} \\
\midrule
Knowledge (how-do-I)   & 58\% (30/52) & 65\% (34/52) \\
Reconfigure            & 17\% (5/30)  & \textbf{93\% (28/30)} \\
Overall ($n{=}82$)     & 43\% (35/82) & \textbf{76\% (62/82)} \\
\bottomrule
\end{tabular}
\caption{Self-knowledge and self-reconfiguration (\code{claude-haiku-4-5}, judged by
\code{claude-sonnet-5}). Docs memory helps knowledge modestly but is decisive on
reconfiguration, where retrieved knob descriptions supply the exact file, name, and
default. The docs condition also adds an explain-don't-act instruction;
Appendix~\ref{app:examples} separates its effect from that of the retrieved
identifiers.}
\label{tab:axisB}
\end{table}
