\section{Related Work}
\label{sec:related}

\paragraph{Natural language to commands.}
NL2Bash \citep{lin2018nl2bash} and Tellina \citep{lin2017tellina} established the
NL-to-shell task and its corpora; we build on the task rather than the models,
personalizing translation with the user's own history.

\paragraph{Retrieval and memory.}
Our memory is retrieval-augmented generation \citep{lewis2020rag} at its lexical
floor: the neural line retrieves with learned dense encoders
\citep{guu2020realm,karpukhin2020dpr}, fuses passages in the decoder
\citep{izacard2021fid}, retrieves per token \citep{khandelwal2020knnlm}, and learns
when to retrieve \citep{asai2024selfrag}; we hold at BM25 \citep{robertson2009bm25}:
the terminal's lexically stable vocabulary makes dense retrieval's paraphrase
advantage marginal (\S\ref{sec:whysimple}). Agent-memory systems page context like an
OS \citep{packer2023memgpt}, distill experience into retrievable reflections
\citep{park2023generative}, and decay memories on an Ebbinghaus schedule
\citep{zhong2024memorybank}; temporal reasoning over such stores remains hard
\citep{maharana2024locomo}. The keybinding failures of
\S\ref{sec:eval}, where a parametric prior overrides a
retrieved fact, are terminal-native instances of the knowledge conflicts studied in QA
\citep{longpre2021entity}; the personalization axis relates to LaMP
\citep{salemi2024lamp}, though our profile is captured from use rather than provided.

\paragraph{Tool use and acting.}
Directive-style acting relates to ReAct \citep{yao2023react}, tool learning to
Toolformer \citep{schick2023toolformer}, and grounding calls against large API
surfaces to Gorilla \citep{patil2023gorilla} and ToolLLM \citep{qin2024toolllm}; we use a
portable text protocol validated against a live registry rather than provider-native
tool calls.

\paragraph{Computer-use and web agents.}
A parallel line of work drives whole desktops and browsers rather than terminals:
computer use \citep{anthropic2024computeruse} operates GUIs from screenshots, OSWorld
\citep{xie2024osworld} benchmarks such agents, OS-Atlas \citep{wu2025osatlas}
pretrains a GUI-grounding action model, and Plan-and-Act \citep{erdogan2025planandact}
separates planning from execution to survive long horizons. Closest to our memory
thesis, WebATLAS \citep{cheng2025webatlas} equips a web agent with experience-driven
memory and simulated action rollouts; MARS brings the same experience-driven memory
to the terminal, where grounding is a registry lookup rather than a vision problem.

\paragraph{Skill induction.}
Voyager grows an open-ended library of verified skills \citep{wang2023voyager}, and
Agent Workflow Memory induces reusable workflows from past trajectories
\citep{wang2024awm}; \S\ref{sec:future} adapts this recipe to the terminal, mining
procedures from the reflex store.
