\section*{Limitations}

Four limitations bound these results. Firstly, the gold sets, though frozen before any
run, are modest (112 and 82 items), so the personalized results should be read as
evidence that memory closes this gap on this set, not as a claim at scale. Secondly,
only one model is under test, so the numbers characterize the small, inexpensive regime
MARS targets rather than models broadly; whether paraphrase generalization holds for a
larger or reasoning model is an open question. Thirdly, the judge (claude-sonnet-5) is
stronger than the model under test but comes from the same model family, so a
family-level preference cannot be ruled out, and an LLM judge carries noise in any case;
the appendix quantifies the observed slack at a few points in either direction. Finally,
the corrective-memory store contained only the 24 seeded pairs during the experiment. A
long-lived store holds hundreds of entries, and although the terminal domain's lexical
stability favors BM25 (\S\ref{sec:whysimple}), we have not measured retrieval precision
against a realistic population of distractors.

\section*{Ethics Statement}

MARS reads the user's shell history (\code{\$HISTFILE}) and stores accepted commands in
a plaintext local file (\code{\char`\~/.mars/cmd\_memory}). Shell history can contain
secrets: tokens pasted into commands, passwords embedded in URLs, internal hostnames.
Retrieval could therefore inject such strings into a prompt bound for a cloud provider.
Every memory and history line is passed through a redaction pass before prompt
injection (known credential prefixes, \code{password=}/\code{token:}-style values, URL
credentials, plus a user-editable denylist at \code{\char`\~/.mars/denylist}); the store
itself stays intact and user-editable, in-editor actions expose, edit, and erase it, and
users who work with sensitive history can point MARS at a local model, which keeps
prompts on the machine entirely. Redaction is heuristic and cannot catch every secret
shape. On the credential side, the key broker
(\S\ref{sec:system}) is itself a mitigation: it removes the incentive to scatter API
keys across remote machines. The evaluation involved no human subjects and no data
beyond author-written gold sets and the authors' own usage logs.
