\section{Qualitative Examples and Error Analysis}
\label{app:examples}

Every example in this appendix is quoted verbatim from the frozen result files the
evaluation was scored on: \code{axis\_a.jsonl} (112 translation items),
\code{lift.jsonl} (24 leakage-controlled cold/paraphrase/verbatim items), and
\code{axis\_b.jsonl} (self-knowledge, 164 rows over the two memory conditions), with the
model under test \code{claude-haiku-4-5} and the judge \code{claude-sonnet-5}
throughout. We organize
the analysis by experiment, and within each by impressive behavior, unexpected errors,
phenomena worth naming, and concrete recommendations. The purpose is to show, from the
raw data, exactly where a simple local-memory mechanism helps, where it does not, and
where the measurement itself is the limiting factor.

\subsection{Axis A: NL$\rightarrow$shell translation}

\paragraph{Impressive behavior.}
On general commands the small model is a competent shell author, and its correct outputs
are not only the easy one-liners: it composes multi-stage pipelines with the right
utility at each stage, including summation and column projection in \code{awk}, buffered
streaming, and process substitution. These were all judged correct:

\begin{footnotesize}
\begin{verbatim}
a20  find the ten largest files under the cwd
   -> find . -type f -printf '%s %p\n' \
        | sort -rn | head -10
a66  count total lines across all python files
   -> find . -name "*.py" -type f -exec wc -l {} + \
        | tail -1
a07  kill the process listening on port 8080
   -> lsof -ti:8080 | xargs kill -9
a30  count how many times each unique line appears
   -> sort | uniq -c
a80  print columns 1 and 3 of a comma-separated file
   -> awk -F',' '{print $1,$3}' filename
\end{verbatim}
\end{footnotesize}

\noindent This is the premise of the paper made concrete: general shell translation is a
near-solved commodity even on a small, inexpensive model (79 of 88 general items
correct), so the accuracy that matters is the personalized slice, which the model cannot
know from the request text alone.

\paragraph{Unexpected errors.}
Two error classes account for the general-command failures, and, unlike the free-tier
runs of \S\ref{sec:setup}, this run produced zero empty completions, so every failure
is a real answer rather than a token-wall blank. The first class is \emph{placeholder and
precision misses}: the model guesses a stand-in filename that the gold set spells
differently (\code{a69} ``print the contents of a file with line numbers''
$\rightarrow$ \code{cat -n}, gold \code{cat -n file.txt}; \code{a31}
\code{grep error log file} vs.\ \code{grep error app.log}), or picks a plausible but
semantically different tool (\code{a38} ``show my current IP'' $\rightarrow$
\code{ip addr show}, a LAN address, where the reference wants \code{curl ifconfig.me};
\code{a86} reads from \code{/dev/stdin} where the gold splits a file into words). The
second class dominates the personalized slice: \emph{ecosystem blindness}, in which the
model answers a Rust or \code{uv} project with the statistically commonest ecosystem:
\code{p04} ``run the unit tests'' $\rightarrow$ \code{npm test} (gold \code{cargo test}),
\code{p06} ``lint the code'' $\rightarrow$ \code{eslint .} (gold \code{ruff check .}),
\code{p11} ``build the release binary'' $\rightarrow$ \code{make release} (gold
\code{cargo build -{}-release}), \code{p18} ``install project dependencies'' $\rightarrow$
\code{npm install} (gold \code{uv sync}). A milder personalized pattern is \emph{honest
abstention}: on the hardest project commands the model declines rather than bluff
(\code{p03} ``deploy the model,'' \code{p12} ``evaluate the latest checkpoint''
$\rightarrow$ ``I need more context\dots''), which is safe behavior but still scored
wrong against a gold command. Both classes are unknowable from the request text, and both
are exactly what corrective memory targets.

\paragraph{Phenomena.}
Even under the stronger \code{claude-sonnet-5} oracle, the judge assigns a \emph{partial} band
that captures near-misses, and reading it is informative: the general failures split into
6 outright wrong and 3 partial, and the partials are precisely the placeholder-filename
cases (\code{a22} \code{chmod +x script} vs.\ gold \code{script.sh}; \code{a77}
\code{zip -r archive.zip .} vs.\ \code{\dots folder/}) where the model understood the
task but guessed a different stand-in. The same slack explains why the two no-memory
measurements of the 24 personalized items disagree: the base row of
Table~\ref{tab:axisA} (42\%, 10/24) and the lift experiment's cold condition (38\%,
9/24) come from separate runs, and they differ on five items. Three differ because the
cold completions were regenerated and the model produced a different command
(\code{p10}, \code{p14}, \code{p17}); two differ because the judge graded the identical
output differently across runs (\code{p01} \code{source venv/bin/activate}, correct
vs.\ partial; \code{p22} \code{docker push}, wrong vs.\ correct). The disagreements move
in both directions and net to one item. The practical consequence is that every point
estimate in this paper carries a couple of points of generation-and-judging slack, and
we read the results through that band.

\paragraph{Recommendations.}
Two inexpensive fixes would tighten the measurement without changing the system's
thesis. Firstly, placeholder-normalized scoring, or explicit partial credit, so that a
filename stand-in is not counted as a comprehension error; the 6 cold \emph{partials} of
the lift set (below) show how much signal lives in this band. Secondly, marking honest
abstentions distinctly from wrong guesses, since ``I need more context'' on a
personalized command is a different failure mode from confidently emitting the wrong
ecosystem. Neither fix touches the model or the memory mechanism; both are evaluation
hygiene.

\subsection{Corrective memory}

\paragraph{Impressive behavior.}
This is the headline result, and the per-item data shows the mechanism cleanly, under the
leakage-controlled protocol in which the store holds only a \emph{paraphrase} of each
prior request (mean word-Jaccard 0.36 to the test query). With those reworded exemplars
retrieved by BM25 into the prompt, 23 of 24 project-specific items are correct, up from 9
correct cold: 15 items flip from wrong-or-partial to correct, and one regresses (below).
Table~\ref{tab:lift-examples} quotes representative flips verbatim. Two sub-mechanisms
are visible. The first is \emph{ecosystem correction}: a Rust or \code{uv}-native project
answered with the commonest default (\code{npm}, \code{eslint}, \code{shellcheck}) is
corrected to the user's real toolchain. The second is \emph{exact-argument recall}: the
model already had the right utility but the wrong specifics, and memory supplies the
precise config, path, or split the user expects reproduced (\code{p02} ``run the training
job'' gains \code{-{}-config configs/base.yaml}; \code{p14} ``run inference on the test
set'' becomes \code{python infer.py -{}-split test}).

\begin{table*}[t]
\centering
\footnotesize
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Request} & \textbf{Cold (no memory)} & \textbf{+\,Paraphrase memory} \\
\midrule
run the unit tests             & \texttt{npm test}                & \texttt{cargo test} \\
lint the code                  & \texttt{eslint .}                & \texttt{ruff check .} \\
build the release binary       & \texttt{make release}            & \texttt{cargo build -{}-release} \\
install project dependencies   & \texttt{npm install}             & \texttt{uv sync} \\
check code formatting          & \texttt{shellcheck *.sh}         & \texttt{black -{}-check .} \\
clean the build artifacts      & \texttt{rm -rf build/ dist/ \dots} & \texttt{cargo clean} \\
run the training job           & \texttt{python train.py}         & \texttt{python train.py -{}-config configs/base.yaml} \\
deploy the model               & \textit{(asks for more context)} & \texttt{./scripts/deploy.sh -{}-prod} \\
\bottomrule
\end{tabular}
\caption{Representative wrong/partial$\rightarrow$correct flips from the
leakage-controlled experiment (\code{lift.jsonl}). The store held only a
\emph{paraphrase} of each prior request (mean word-Jaccard 0.36) and the model was tested
on the original, so a warm win requires generalization across the paraphrase gap, not
lookup. The top block shows \emph{ecosystem correction} (a Rust/\code{uv} project
answered with \code{npm}/\code{eslint}/\code{shellcheck} defaults); the last two rows
show \emph{exact-argument recall} and \emph{abstention rescue}. All warm outputs match
gold.}
\label{tab:lift-examples}
\end{table*}

\paragraph{Unexpected errors.}
The paraphrase column has a single error out of 24, and it is also the experiment's one
\emph{regression}: an item that was correct cold became wrong with memory, which is why
9 cold-correct plus 15 flips yields 23 warm-correct rather than 24. Cold, \code{p19}
``run the benchmark suite'' was answered \code{make benchmark}, which the judge accepted
as a plausible route to the gold \code{python -m benchmarks.run}. With the paraphrased
exemplar retrieved, the model instead answered \code{cargo bench}: it resolved the
ecosystem to Rust, and the retriever did not surface the paraphrased Python exemplar
strongly enough to override that prior. This is a retrieval-plus-prior miss, not a
lookup failure, and it is the honest residual of a method that must generalize rather
than copy. The shape of the cold column is also informative: of the 15 non-correct
items, 9 are outright wrong and 6 partial, so memory is repairing a mix of
fundamentally-wrong ecosystem guesses and last-mile precision misses, not merely
polishing near-correct commands.

\paragraph{Phenomena.}
The gap between the two warm conditions is the cleanest measure of what BM25 buys here.
Verbatim memory (the exact prior request in the store) scores 100\% (24/24), the trivial
lookup ceiling, while paraphrase memory scores 96\% (23/24). The paraphrase condition
falls only one item short of verbatim lookup despite a mean word-Jaccard of just 0.36
between the stored exemplar and the query: the retriever bridges a substantial rewording
gap on 23 of 24 items, so the personalization win is a genuine generalization result
rather than an artifact of a planted answer. The one-item gap is the price of that
generalization.

\paragraph{Recommendations.}
The residual is a \emph{retrieval} miss, not a generation miss: when the paraphrase
drifts far enough and a strong ecosystem prior competes, lexical BM25 does not surface
the stored pair decisively. The cheapest next lever, given that stored commands now carry
\code{ts}/\code{session}/\code{cwd} fields, is recency-and-project weighting: when
several exemplars tie lexically, prefer the user's recent and same-directory commands.
Semantic retrieval with embeddings is the rung after that, and is warranted only once the
domain broadens enough that the paraphrase gap dominates
(\S\ref{sec:future}).

\paragraph{A note on token cost.}
Memory's effect on token cost is model-dependent, an observation that cuts against the claim
we expected to make. On non-reasoning \code{claude-haiku-4-5}, retrieved
exemplars add a small input cost (mean 95$\rightarrow$161 tokens per call, $+66$). On a
\emph{reasoning} model (Qwen3-32B, from the free-tier runs of \S\ref{sec:setup}),
however, the same few-shot grounding suppresses the verbose chain-of-thought the model
otherwise emits when unsure, so net cost falls (385$\rightarrow$217 per call, $-44\%$).
Memory as a token \emph{reduction} is therefore a reasoning-model phenomenon, not a
universal property; we report the per-model effect rather than headline the favorable
case.

\subsection{Axis B: self-knowledge}

Docs memory lifts overall accuracy from 43\% to 76\% (35$\rightarrow$62 of 82), but the
two question types behave very differently, and the split between them is informative.
Retrieval turns a plausible guess into a grounded fact:

\begin{footnotesize}
\begin{verbatim}
k02 "How do I save the current file?"
 none: "Use the Save action."
       [would run: Save]              (acts, no key)
 docs: "Press C-x C-s to save
        the current buffer."          (correct)
r16 "How do I change the accent color?"
 none: "Mars doesn't have built-in
        theme color settings ..."     (wrong)
 docs: set theme_accent to an [r,g,b]
       array in tuning.json           (correct)
\end{verbatim}
\end{footnotesize}

\noindent The reconfigure slice is where memory is decisive, rising from 17\% to 93\%
(5$\rightarrow$28 of 30). Without docs the model invents plausible config locations:
\code{r16} ``change the accent color'' is answered ``Mars doesn't have built-in theme
color settings,'' and \code{r04} ``use a different model'' points at a
\code{\char`\~/.mars/config.json} that does not exist. With the retrieved knob descriptions,
which carry the exact name, file, and default, the agent answers \code{theme\_accent} /
\code{MARS\_LLM\_MODEL} in \code{tuning.json} correctly. Two failure families that
motivated the corpus design, \emph{acts-instead-of-explains} (answering a how-to question
with a bare \code{[would run: \dots]} directive, \code{k02}, \code{k09}) and
\emph{hallucinated identifiers} (a right-idea knob under the wrong filename), are largely
eliminated on this slice by the actionable, path-carrying knob descriptions and an
explain-don't-act instruction on the docs path.

Knowledge questions gain far less, from 58\% to 65\%, for an instructive reason: they
turn on an \emph{exact keybinding}, and even with the docs retrieved the model sometimes
prefers a plausible chord over the one the docs state (\code{k05} ``detach my session''
$\rightarrow$ \code{C-x C-c}, gold \code{C-x C-d}; \code{k06} ``search within a buffer''
$\rightarrow$ \code{C-f}, gold \code{C-s}; \code{k18} ``jump to top/bottom''
$\rightarrow$ \code{C-x Home}, gold \code{M-<}). A knob \emph{name} is a token the
retriever can hand over intact; a chord is a specific fact the model will still overwrite
with a more common default, which is why doc memory helps reconfiguration far more than
keybinding recall. A residual limit is judge slack: a few correct-but-differently-phrased
answers are marked wrong, so the reported knowledge number is a mild lower bound even
under the stronger \code{claude-sonnet-5} oracle.
