\documentclass[11pt]{article}

% "preprint" gives a non-anonymous version with page numbers (EMNLP system
% demonstrations are not anonymized). Switch to "final" for camera-ready.
\usepackage[preprint]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,shapes.geometric}

\newcommand{\code}[1]{\texttt{#1}}

\title{MARS: A Memory-Augmented Agentic Terminal Multiplexer\\ that Learns From Your Commands}

\author{Anjishnu Kumar\thanks{~Work done before joining Microsoft.} \\
  Microsoft Copilot AI \\
  Redmond, WA \\
  \texttt{anjikumar@microsoft.com} \\}

\begin{document}
\maketitle

\begin{abstract}
The LLM assistants a developer uses at the terminal are generic and amnesiac: they run
in a separate window, see only what is pasted into them, forget the project-specific
command you corrected yesterday, and --- on a remote GPU box --- require an API key that
leaks into that box's environment and shell history. We present \textbf{MARS}, a
single-binary terminal editor, multiplexer, and built-in agent that closes this gap with
\emph{local memory}: because the agent is embedded where the user works, it retrieves two
kinds of context a cloud model cannot reach --- the user's own past commands, and the
tool's own documentation --- and injects them into the prompt. The retrieval is
deliberately trivial: lexical BM25 over a local file, no embeddings, nothing that leaves
the machine. In a constrained domain this simple mechanism has an outsized effect. On personalized commands, feeding a small, inexpensive
model (claude-haiku-4-5) a few \emph{paraphrases} of the user's own past corrections
raises translation accuracy from \textbf{37\% to 95\%} under a leakage-controlled
protocol --- the store holds only reworded prior requests, so a win requires retrieval to
bridge a paraphrase gap, not look up a planted answer. Pointed at the tool's own
documentation, the same mechanism lifts self-knowledge accuracy from \textbf{42\% to 75\%}
(and mapping a plain-English request to the right configuration knob from \textbf{16\% to
93\%}). We describe the agent's minimal machinery (a registry-validated directive protocol,
a corrective-memory loop, a free-model cascade), the architecture that makes always-there
local memory possible (a persistent daemon and a key-never-leaves-home SSH broker), and a
self-instrumenting evaluation in which the agent's debug log is the benchmark corpus.
\end{abstract}

\section{Introduction}

Consider an ML researcher training models across a laptop and a handful of remote GPU
boxes. Over a day they babysit long runs in a shell, edit configs and model code in an
editor, hold several sessions open with a multiplexer, and ask an LLM to explain a CUDA
out-of-memory trace or draft an \code{rsync} command. The assistant is the weakest link,
for two reasons that compound. It is \textbf{generic}: living in a separate window, it sees
only the fragment the user pastes in, and does not know that in \emph{this} project a
training run is launched with a particular \code{sbatch} incantation settled on last week.
And it is \textbf{amnesiac}: it forgets that command between sessions, forgets ``where was
I'' on reattaching to an overnight run, and knows nothing about the editor it is embedded
in. Worse, in the remote setting that defines GPU work, reaching the work at all means
exporting an API key onto a box the researcher does not own, where it lingers in the
environment, shell history, and often a dotfile.

Our insight is that an agent \emph{embedded at the terminal} has access to memory a cloud
model structurally cannot reach. Two local corpora sit within arm's length of a
terminal-resident agent and nowhere near a browser tab: the user's own history of commands
they actually ran, and the tool's own documentation, action registry, and configuration
schema. Neither is large, secret to the machine, or requires training --- they only require
an agent that lives where they live. We built \textbf{MARS} (a Memory-Augmented Rust Shell)
to exploit exactly this: it retrieves from both corpora with plain lexical search and
injects the results into the prompt.

The surprise is how little machinery this takes and how much it buys. We expected local
memory to help; we did not expect a trivial lexical method to nearly close the gap between
a small, inexpensive model and its personalized ceiling --- and to do so even when the
stored memory is only a \emph{paraphrase} of the user's earlier request rather than its
verbatim text. On a set of project-specific commands, retrieving a handful of reworded
past corrections lifts claude-haiku-4-5 from 37\% to 95\% accuracy, with the retriever
forced to bridge the paraphrase gap rather than echo a planted answer. In a narrow domain,
memory need not be sophisticated to be decisive.

This paper makes four contributions. First, a memory-augmented terminal agent (\S2) built
from minimal parts --- a registry-validated directive protocol, a corrective-memory loop,
and a free-model cascade. Second, two axes of \emph{local} memory: the user's own commands
for translation, and the tool's own ground truth for self-knowledge and reconfiguration.
Third, the architecture (\S3)
that makes always-there local memory possible: a persistent daemon and a
key-never-leaves-home SSH broker that keeps the credential off remote boxes. Fourth, a
self-instrumenting evaluation (\S4) in which the agent's debug log doubles as the
benchmark corpus, and in which corrective memory takes a small model from 37\% to 95\% on
personalized commands under a leakage-controlled protocol. MARS is open source (MIT),
installs with
\code{cargo install mars-terminal}, and is about 9{,}500 lines of Rust.

\section{The Memory-Augmented Agent}

\begin{figure*}[t]
\centering
\begin{tikzpicture}[
  scale=0.79, transform shape,
  font=\footnotesize,
  >={Latex[length=2mm]},
  term/.style={draw, rounded corners, align=center, minimum width=23mm, minimum height=11mm, fill=black!4},
  core/.style={draw, very thick, rounded corners, align=center, minimum width=26mm, minimum height=11mm, fill=black!8},
  mem/.style={draw, cylinder, shape border rotate=90, aspect=0.16, align=center,
              minimum width=27mm, minimum height=11mm, fill=black!3, inner sep=1pt},
  llm/.style={draw, dashed, rounded corners, align=center, minimum width=25mm, minimum height=12mm, fill=black!2},
  note/.style={font=\scriptsize\itshape, align=center, text=black!65},
  el/.style={font=\scriptsize, align=center},
]

% --- nodes ---
\node[term] (term) at (0,0) {\textbf{Terminal}\\ \textbf{multiplexer}\\[1pt] \scriptsize editor + shell panes};
\node[core] (core) at (5.4,0) {\textbf{Agent}\\ \textbf{core}};
\node[mem]  (m1) at (3.35,-2.35) {\textbf{M1: your commands}\\[1pt] \scriptsize \code{\char`~/.mars/cmd\_memory}\\ \scriptsize + shell history};
\node[mem]  (m2) at (7.45,-2.35) {\textbf{M2: tool self-knowledge}\\[1pt] \scriptsize docs, action registry,\\ \scriptsize tuning knobs};
\node[llm]  (llm) at (12.7,0) {\textbf{LLM}\\ free-model\\ cascade};
\node[note] (log) at (5.4,1.55) {self-instrumenting log\\ $=$ benchmark corpus};

% --- your-machine boundary (behind) ---
\begin{scope}[on background layer]
  \node[draw, dashed, thick, rounded corners=6pt, fill=blue!3, inner sep=10pt,
        fit=(term)(core)(m1)(m2)(log)] (mach) {};
\end{scope}
\node[anchor=north west, font=\scriptsize\bfseries, text=blue!45] at (mach.north west) {~your machine};

% --- terminal <-> core ---
\draw[->] ([yshift=2.4mm]term.east) -- ([yshift=2.4mm]core.west)
      node[midway, above, el] {NL request / question};
\draw[<-] ([yshift=-2.4mm]term.east) -- ([yshift=-2.4mm]core.west)
      node[midway, below, el] {directive:\\ \code{RUN}/\code{TYPE}/\code{OPEN}\\ (registry-validated)};
\draw[->, dashed] (term.north) to[bend left=22] node[midway, above, note] {sees the panes (sight)} (core.north west);

% --- core <-> memory (BM25) ---
\draw[<->] (core.south) -- (m1.north);
\draw[<->] (core.south) -- (m2.north);
\node[el] at (5.4,-1.35) {BM25 retrieve\\ $\rightarrow$ few-shot into prompt};

% --- core <-> LLM (the only thing that leaves) ---
\draw[->, very thick] ([yshift=2.4mm]core.east) -- ([yshift=2.4mm]llm.west)
      node[midway, above, el] {prompt};
\draw[<-, very thick] ([yshift=-2.4mm]core.east) -- ([yshift=-2.4mm]llm.west)
      node[midway, below, el] {completion};
\node[note, text=red!55] at (10.05,-1.05) {the only call that\\ leaves your machine};

% --- broker annotation ---
\node[note] (brk) at (12.7,-2.2) {\textbf{key-never-leaves-home}\\ on a remote box the call is\\ proxied home; the key stays local};
\draw[->, dotted, black!55] (brk.north) -- (llm.south);

\end{tikzpicture}
\caption{MARS augments a terminal LLM agent with \emph{local memory over the user's own
context}. Seeing the multiplexer's panes, the agent retrieves with lexical BM25 from two local
stores --- \textbf{M1}, the user's past commands, and \textbf{M2}, the tool's own docs and
registry --- and injects the results as few-shot exemplars. Only the model call leaves the
machine (on a remote box even the key is proxied home), and every call is logged into the
corpus the evaluation is built from. The dashed boundary is the thesis: memory, keys, and
benchmark all stay local.}
\label{fig:arch}
\end{figure*}

\subsection{Tasteful minimalism}

The recurring risk in an agentic editor is that the agent becomes a second system bolted
beside the first, with its own command language, state, and failure modes. Our design
principle is the opposite: add the \emph{minimum} machinery that makes memory augmentation
work, and reuse everything else. The agent is not a subsystem but another way of naming an
action. Everything MARS can do is a variant of one \code{Action} enum, and three interfaces
resolve a user's intent to it --- a direct keychord (procedural recall), a fuzzy command bar
(recognition), and the agent (natural language) --- all terminating in a single dispatch
point. A capability added once is therefore chord-bindable, searchable, and agent-invokable
for free (Figure~\ref{fig:arch} sketches the retrieval loop).

This trades expressive freedom for grounding. Because the agent can only name an action
that exists in the registry, and retrieves from flat local files rather than a vector
index, the surface a user must trust is small and legible. We pay for this in ceiling: MARS
cannot invent a capability at runtime, and lexical retrieval will miss a paraphrase an
embedding would catch (\S4). In the terminal domain the trade is strongly favorable, and the
rest of this section inventories the few parts we did add.

\subsection{The agent as a validated input device}

An agent that can act on the user's workspace must be grounded so it cannot act on a
capability that does not exist. The principle, following the mode-error tradition in
interface design, is that the agent is one input device among several and must be
\emph{safer} than a keypress, never a bypass around the editor's gates. The implementation
is a trailing-line directive protocol: the model receives a live-generated catalog of every
action, its description, and its keybinding, and ends its reply with at most one directive
(Table~\ref{tab:directives}). We chose a text protocol over provider-native tool-calling APIs
deliberately --- it works identically across OpenAI-compatible endpoints, a native Anthropic
path, and local open models, and renders as human-readable text the user confirms before
anything fires --- at the cost of the stricter typing a schema gives.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Directive} & \textbf{Effect (gated by confirm)} \\
\midrule
\code{RUN: <Action>}   & fire a registry action \\
\code{TYPE: <cmd>}     & type a command into a shell pane \\
\code{OPEN: path:line} & jump to a file location \\
\code{NEED: <context>} & pull more context; re-ask once \\
\bottomrule
\end{tabular}
\caption{The directive protocol. A \code{RUN:} naming an action absent from the live
registry is rejected before it fires, so a hallucinated action cannot execute; a
destructive action passes the same confirmation gate a human would.}
\label{tab:directives}
\end{table}

The grounding is the registry check. A \code{RUN:} whose name does not resolve against the
live \code{Action} set is rejected before dispatch, so the model cannot execute a
hallucinated capability --- the same enum that defines what MARS can do defines what the
agent may invoke. \code{NEED:} closes a different gap: rather than always dumping the full
screen, the model can request more context (a pane's scrollback, another tab) on demand, and
MARS supplies it and re-asks exactly once, the depth hard-capped so no retrieval loop forms.
This is a small, model-driven form of the retrieval this paper is about, applied to the live
screen rather than to memory.

\subsection{Natural-language shell translation}

The most frequent agentic request at a terminal is ``turn this English into the right
command.'' The principle is to keep the human in the loop cheaply: translation should be
assistive and reversible, never an automatic execution. The implementation lives on the
command bar --- \code{Ctrl+Space}, then plain English (``find the biggest checkpoints
here'') --- which sends the request with the on-screen context (working directory, recent
output) to the model cascade (\S2.6) and renders the proposed command at the cursor to run
or discard. Translation alone is commodity capability
\citep{lin2018nl2bash,lin2017tellina}; what makes it work \emph{for a given user} is the
memory we turn to next.

\subsection{Memory 1: the user's own commands}

A generic model does not know your commands. Asked to ``launch the training run,'' it
produces a plausible \code{python train.py} when this project in fact uses
\code{sbatch --gres=gpu:8 scripts/train.slurm} --- a command the user has typed a dozen
times and the model has never seen. The principle is that the user's own accepted commands
are the highest-value context available, and are local to the terminal by construction. The
implementation is a corrective loop. Every time the user accepts (or edits and runs) a
translated command, MARS appends the \code{(request, command)} pair to a local JSON-lines
file, and also reads recent shell history from \code{\$HISTFILE}. On the next request, MARS
ranks these entries against the query with BM25 \citep{robertson2009bm25} --- $k_1{=}1.5$,
$b{=}0.75$, no index to build --- and injects the top few as few-shot exemplars.

The loop is \emph{corrective} in the precise sense that a command the model got wrong once
becomes one it gets right thereafter, because the user's correction is what enters the store
--- a capability a stateless cloud assistant cannot have, since it never observed the
correction and does not persist between calls. Nothing here is novel as information retrieval;
the contribution is that in this domain the retrieval can be this trivial and still be
decisive (\S4). The cost is a bounded prompt inflation --- a few exemplars added to the input,
which \S4 measures and finds far outweighed by the accuracy
gain (and on a reasoning model the grounding can even shorten the output enough to pay it
back).

\subsection{Memory 2: the tool's own documentation}

An agent embedded in a tool should answer questions about that tool, and change its settings,
from ground truth rather than a guess. The principle is that MARS's own capability surface ---
the action registry, keybindings, and tuning knobs, each carrying a human-written description
--- is authoritative documentation generated from the code, and cannot drift. The
implementation reuses the same BM25 retriever over a second corpus: MARS's documentation
chunked into passages, plus the live registry and the described knobs. On a \code{NEED: docs}
pull, MARS injects the top passages, so ``how do I split the screen?'' is answered with the
real keybinding and ``make the accent orange'' resolves to the actual knob and a confirmable
edit. When retrieval returns nothing, the agent says the capability does not exist rather than
invent one --- honest refusal grounded in the same registry that gates \code{RUN:}.

\subsection{Why simple memory is enough}

It is worth stating why a method this trivial suffices, because the instinct in a memory
system is to reach for embeddings and a vector database. The terminal is a \emph{constrained}
domain. A user's command vocabulary is small, repetitive, and lexically stable --- the same
flags, paths, and tool names recur verbatim --- so query and useful memory share surface
tokens, exactly the regime in which BM25 is strong and dense retrieval's paraphrase advantage
is marginal. The corpora are small (hundreds of entries), so no scaling pressure justifies an
index. And the payoff is asymmetric: a single relevant exemplar of the user's own command is
worth more than semantic breadth, because it supplies the exact string the user wants
reproduced. We therefore treat embeddings as a cost to be deferred until the domain broadens
(\S6). The empirical case for this choice is the
37\%-to-95\% paraphrase-memory result of \S4 --- where BM25 bridges a mean word-Jaccard
gap of 0.36 between the query and the stored exemplar --- and the mechanism producing it is
a prompt, a local file, and a standard ranking function.

\subsection{The model cascade}

Most terminal LLM traffic is cheap and can run on free models, but a fixed choice of one
hits rate limits and wastes quality where it is not needed. The principle is to right-size
per task, distinguishing two orthogonal reasons to change models. Rotating \emph{for limits}
moves a task horizontally across same-tier open models on a rate-limit error: because each
family is metered on its own counter, rotating across $N$ families multiplies the effective
free ceiling by roughly $N$, never lowering quality because it stays in-tier. Escalating
\emph{for quality} moves a task vertically to a stronger tier, only on a \emph{detected}
failure --- which MARS gets for free, because its output is structured: a \code{RUN:} that
fails the registry check is an unambiguous signal a small model erred, triggering one retry at
a higher tier. This trades a little dispatch complexity for the free-tier headroom the token
wall of \S4.4 makes concrete.

\section{System: The Substrate for Local Memory}

Local memory is only useful if the agent is reliably present where the memory lives, across
disconnects and on machines the user does not own. The architecture exists to guarantee that
presence, and is kept as small as the agent.

\paragraph{One binary, one core.}
MARS is a single Rust binary. Its application core owns all state --- buffers, panes,
terminals, the agent conversation, the memory stores --- and never touches a TTY directly:
input arrives as source-neutral events, and rendering is a stateless projection onto a
\code{ratatui} backend. The same core therefore runs unmodified in three configurations ---
a standalone terminal, a headless daemon, and a test backend --- which lets the evaluation
drive the real system with no mocks (\S4).

\paragraph{A persistent daemon.}
Persistence is a process split, not a save-and-restore feature. A server process runs the
whole application headless and emits the editor's own ANSI byte stream over a Unix socket; a
thin client owns the real terminal and pumps bytes. We chose to have the server render ANSI
rather than define a structured protocol because it reuses the entire existing rendering
pipeline with no second renderer to maintain; the cost is that the wire carries rendered
frames rather than semantic events. The payoff for memory is direct: because the core keeps
running with no client attached, the command store and the user's session survive a closed
laptop, and a run left going overnight is watched and summarized for the user's return.

\paragraph{Classifier-free routing.}
The command bar routes by \emph{sigil}, not by a learned classifier: the first character
names the target namespace --- \code{!} runs a shell command, \code{?} asks the agent,
\code{@} opens the file navigator, and plain text fuzzy-searches actions. Routing is
deterministic and never mispredicts, so there is no routing model to train, evaluate, or
explain, and no classification error for the user to work around.

\paragraph{The key never leaves home.}
The remote-work deficit is a security problem: exporting a key onto every GPU box scatters the
credential across machines the user does not control. Our principle is that a secret is a fact
about the user, not the machine, so it should live in one place. The implementation is a
broker. A home-side daemon (\code{mars keyd}) holds the key; when the user runs \code{mars
ssh}, MARS reverse-forwards the broker's socket over the SSH connection, and the agent on the
remote box makes its model calls by proxying the request \emph{home}, where the broker injects
the credential and streams the completion back. The key never lands on the remote --- not on
disk, in the environment, or in memory --- so a compromised box has nothing to steal, and
access ends when the tunnel closes.

\paragraph{The log is the benchmark.}
Running with \code{--llm-debug} appends one JSON line per model call --- task, model, real
token counts, latency, and the full prompt and reply --- to a local stream, and records
whether each suggestion was accepted, edited, or rejected. This log is both a cost profile the
user can inspect and, as \S4 shows, the corpus the evaluation is built from: the system
instruments itself, so the benchmark is drawn from real use rather than authored by hand.

\section{Evaluation}

\subsection{Setup}

We evaluate the two memory axes on frozen, a-priori gold sets and a single small model under
test. The command-translation gold set contains 112 items --- 88 general and 24
project-specific (personalized) --- and the self-knowledge gold set contains 82 question--answer
items (52 knowledge, 30 reconfiguration); both were fixed before any system was run against
them. The model under test is claude-haiku-4-5, a small, inexpensive model: MARS's thesis needs
the model to be \emph{small} --- so personalization has a gap to close --- not necessarily free
(\S4.4 explains why the final run uses a small paid model). To avoid a model grading itself,
correctness is scored by a stronger cross-model judge, claude-sonnet-5 \citep{zheng2023judging}.
Prompts, contexts, and token counts come from the self-instrumenting log of \S3; across all runs
there were 0 empty completions.

\subsection{Axis A: learning the user's commands}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lc@{}}
\toprule
\textbf{Setting} & \textbf{Accuracy} \\
\midrule
General ($n{=}88$)        & 89\% \\
Personalized ($n{=}24$)   & 41\% \\
Overall ($n{=}112$)       & 79\% \\
\midrule
\multicolumn{2}{@{}l}{\emph{Corrective memory (personalized, $n{=}24$, leakage-controlled)}} \\
\quad Cold (no memory)             & 37\% \\
\quad Paraphrase memory            & \textbf{95\%} \\
\quad Verbatim memory (ceiling)    & 100\% \\
\bottomrule
\end{tabular}
\caption{Command translation (claude-haiku-4-5, judged by claude-sonnet-5). The
corrective-memory block is leakage-controlled: the store holds only a \emph{paraphrase} of each
prior request and the model is tested on the original, so paraphrase memory (headline, 95\%)
measures generalization; verbatim memory (100\%) is the trivial lookup ceiling.}
\label{tab:axisA}
\end{table}

The base numbers in Table~\ref{tab:axisA} confirm the premise of the paper. The small model is
already good at general shell translation --- 89\% --- but falls to 41\% on project-specific
commands it cannot know, dragging overall accuracy to 79\%. These are not reasoning failures but
missing-knowledge failures, exactly what local memory is positioned to fix.

The corrective-memory experiment isolates that fix, under a protocol designed to rule out the
objection that memory is mere lookup. For each of the 24 tasks the store is seeded not with the
prior request verbatim but with an independently generated \emph{paraphrase} of it --- a
reworded request with the same intent (mean word-level Jaccard 0.36 to the test query) --- while
the model is tested on the \emph{original} request. A ``warm'' win therefore requires
retrieval to bridge the paraphrase gap and the model to generalize from a
differently-worded exemplar, not to copy a planted answer. Under this setup, cold accuracy is
37\% (9/24); paraphrase memory raises it to
\textbf{95\%} (23/24), with 15 of 24 items flipping from wrong-or-partial to correct; and a
verbatim-memory condition, where the exact prior request is in the store, reaches the trivial
lookup ceiling of 100\% (24/24). The headline is the paraphrase number: simple BM25 over a local
file recovers almost all of the personalization gap by generalizing, not by lookup. Memory adds
a small token cost here (about 66/call), whose sign is model-dependent
(\S4.4, App.~\ref{app:examples}).

\subsection{Axis B: learning the tool itself}

The second axis asks whether doc memory lets the agent answer questions about MARS and
reconfigure it. Injecting retrieved documentation lifts self-knowledge accuracy from 42\% to
75\% overall (Table~\ref{tab:axisB}), but the gain is unevenly distributed. On ``how-do-I''
knowledge questions it is modest --- 57\% to 65\% --- because the no-memory model already
answers many correctly (by naming the right action) and, even with docs, sometimes hallucinates
a wrong keybinding for one the docs state exactly (\S\ref{app:examples}). The decisive gain is on
\emph{reconfiguration} --- ``autosave more
often'' to a concrete knob edit --- rising from 16\% to \textbf{93\%}: without the docs the
model invents config filenames and knob names, while with the tuning-knob descriptions
retrieved it emits the exact \code{knob = value in tuning.json} the user needs.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Question type} & \textbf{No memory} & \textbf{+\,Docs memory} \\
\midrule
Knowledge (how-do-I)   & 57\% & 65\% \\
Reconfigure            & 16\% & \textbf{93\%} \\
Overall ($n{=}82$)     & 42\% & \textbf{75\%} \\
\bottomrule
\end{tabular}
\caption{Self-knowledge and self-reconfiguration (claude-haiku-4-5, judged by claude-sonnet-5).
Docs memory helps knowledge questions modestly but is decisive on reconfiguration, where the
retrieved knob descriptions supply the exact file, name, and default the model otherwise
invents.}
\label{tab:axisB}
\end{table}

\subsection{The free-tier token wall}

The cascade's rationale surfaced as a hard constraint during evaluation itself. Our first
choice for the model under test was a genuinely free one --- but both free tiers we tried,
Qwen3-32B on Groq and Gemini Flash-Lite, exhausted their \emph{daily} token quotas partway
through the roughly 350-call batch, the exact token wall the cascade (\S2.7) routes around by
rotating across metered families. Because a reproducible measurement needs every item scored in one clean pass, we ran the final
numbers on a small \emph{paid} model, claude-haiku-4-5 --- small enough to stand in for the tier
MARS targets (the thesis needs \emph{small}, not free). Tellingly, the free-tier runs emitted
empty completions on hitting the wall, whereas the paid
run had 0 empties --- confirming those blanks were a throughput artifact, not a comprehension
failure. (A model-dependent effect of memory on token cost is in Appendix~\ref{app:examples}.)

\subsection{Limitations}

Three limits bound these results honestly. The gold sets, though frozen and a-priori, are
modest (112 and 82 items), so the personalized results read as ``memory closes this gap on this
set,'' not a claim at scale. Only one model is under test, so the numbers characterize the
small, inexpensive regime MARS targets rather than models broadly; whether paraphrase
generalization holds for a larger or reasoning model is open. And though we judge with a
stronger model than the one under test (claude-sonnet-5 grading claude-haiku-4-5), an LLM judge
still carries noise --- a few points of grading slack either way (\S\ref{app:examples}).

\section{Related Work}

\paragraph{Terminal tooling.}
Multiplexers such as tmux and Zellij give persistence but no intelligence; AI terminals such
as Warp \citep{warp} add command search but see only their own blocks, not the user's editor or
corrections; IDE agents such as Cursor \citep{cursor} run their intelligence on the laptop while
remote state lives on the box, and lose context when the window closes. MARS differs on one
axis: the agent is embedded in a persistent, terminal-resident process, so it holds local memory
none of these architectures host.

\paragraph{Natural language to commands.}
NL2Bash \citep{lin2018nl2bash} and Tellina \citep{lin2017tellina} established the NL-to-shell
task and its corpora; we build on the task rather than the models, personalizing translation
with the user's own history rather than improving the translator in general.

\paragraph{Agents and retrieval.}
Directive-style acting relates to ReAct \citep{yao2023react} and tool use to Toolformer
\citep{schick2023toolformer}, though we deliberately use a portable text protocol with
registry grounding rather than provider-native tool calls. Our memory is retrieval-augmented
generation \citep{lewis2020rag} in its simplest lexical form; the novelty is not the mechanism
but the setting --- retrieval over the user's \emph{own} local context, embedded where that
context is produced.

\section{Conclusion}

A terminal agent embedded where the user works can retrieve two kinds of local memory --- the
user's own commands and the tool's own documentation --- and in this constrained domain BM25
over a local file is decisive: a few reworded exemplars of the user's corrections take a small
model from 37\% to 95\% on personalized commands, and the tool's own docs take it from 16\% to
93\% at reconfiguring itself. We read this as the first rung of a ladder. Above single-command
recall lie \emph{skill learning} (mining repeated command \emph{sequences} into named
procedures), \emph{workflow understanding} (telling a training run from a debug session), and
\emph{cross-fleet memory} (embeddings, once paraphrase matters, following the user across
machines through the same broker that forwards their key). The through-line is ownership: memory
no cloud agent, seeing only what it is handed, can replicate. MARS is open
source~\citep{kumar2026mars}.

\bibliography{refs}

\appendix

\section{Qualitative Examples and Error Analysis}
\label{app:examples}

Every example in this appendix is quoted verbatim from the frozen result files the evaluation
was scored on --- \code{axis\_a.jsonl} (112 translation items), \code{lift.jsonl} (24
leakage-controlled cold/paraphrase/verbatim items), and \code{axis\_b.jsonl} (self-knowledge,
164 rows over the two memory conditions) --- with the model under test claude-haiku-4-5 and the
judge claude-sonnet-5 throughout. We organize the analysis by experiment, and within each by
impressive behavior, unexpected errors, phenomena worth naming, and concrete recommendations.
The point is not to celebrate the good cases but to show, from the raw data, exactly where a
trivial local-memory mechanism helps, where it does not, and where the measurement itself is
the limiting factor.

\subsection{Axis A: NL$\rightarrow$shell translation}

\paragraph{Impressive behavior.}
On general commands the small model is a competent shell author, and its correct outputs are not
only the easy one-liners: it composes multi-stage pipelines with the right utility at each stage,
including summation and column projection in \code{awk}, buffered streaming, and process
substitution. These were all judged correct:

\begin{footnotesize}
\begin{verbatim}
a20  find the ten largest files under the cwd
   -> find . -type f -printf '%s %p\n' \
        | sort -rn | head -10
a66  count total lines across all python files
   -> find . -name "*.py" -type f -exec wc -l {} + \
        | tail -1
a07  kill the process listening on port 8080
   -> lsof -ti:8080 | xargs kill -9
a30  count how many times each unique line appears
   -> sort | uniq -c
a80  print columns 1 and 3 of a comma-separated file
   -> awk -F',' '{print $1,$3}' filename
\end{verbatim}
\end{footnotesize}

\noindent This is the paper's premise made concrete: general shell translation is a near-solved
commodity even on a small, inexpensive model (79 of 88 general items correct), so the accuracy
that matters is the personalized slice the model cannot know from the request text alone.

\paragraph{Unexpected errors.}
Two error classes account for the general-command failures, and --- unlike the free-tier runs of
\S4.4 --- \textbf{this run produced zero empty completions}, so every failure is a real answer
rather than a token-wall blank. First, \textbf{placeholder and precision misses}: the model
guesses a stand-in filename the gold set spells differently (\code{a69} ``print the contents of a
file with line numbers'' $\rightarrow$ \code{cat -n}, gold \code{cat -n file.txt}; \code{a31}
\code{grep error log file} vs.\ \code{grep error app.log}), or picks a plausible but semantically
different tool (\code{a38} ``show my current IP'' $\rightarrow$ \code{ip addr show}, a LAN address,
where the reference wants \code{curl ifconfig.me}; \code{a86} reads from \code{/dev/stdin} where
the gold splits a file into words). Second, on the personalized slice the dominant error is
\textbf{ecosystem blindness} --- the model answers a Rust or \code{uv} project with the
statistically commonest ecosystem: \code{p04} ``run the unit tests'' $\rightarrow$ \code{npm test}
(gold \code{cargo test}), \code{p06} ``lint the code'' $\rightarrow$ \code{eslint .} (gold
\code{ruff check .}), \code{p11} ``build the release binary'' $\rightarrow$ \code{make release}
(gold \code{cargo build --release}), \code{p18} ``install project dependencies'' $\rightarrow$
\code{npm install} (gold \code{uv sync}). A milder personalized pattern is \textbf{honest
abstention}: on the hardest project commands the model declines rather than bluff (\code{p03}
``deploy the model,'' \code{p12} ``evaluate the latest checkpoint'' $\rightarrow$ ``I need more
context\dots''), which is safe but still scored wrong against a gold command. Both are unknowable
from the request text and are exactly what corrective memory targets.

\paragraph{Phenomena.}
Even under the stronger claude-sonnet-5 oracle, the judge assigns a \emph{partial} band that
captures near-misses, and reading it is informative: the general failures split into 6 outright
wrong and 3 partial, and the partials are precisely the placeholder-filename cases (\code{a22}
\code{chmod +x script} vs.\ gold \code{script.sh}; \code{a77} \code{zip -r archive.zip .} vs.\
\code{\dots folder/}) where the model understood the task but guessed a different stand-in. The
cross-run grading instability that plagued earlier drafts is largely gone --- the stronger judge
grades the same cold output consistently across \code{axis\_a.jsonl} and \code{lift.jsonl} --- so
the practical consequence is narrower than before: the Axis-A point estimates should be read with
a couple of points of partial-credit ambiguity, not a large judge-noise band.

\paragraph{Recommendations.}
Two cheap fixes would tighten the measurement without changing the system's thesis. (1)
Placeholder-normalized scoring, or explicit partial credit, so that a filename stand-in is not
counted as a comprehension error --- the 6 cold \emph{partials} of the lift set (below) show how
much signal lives in this band. (2) Marking honest abstentions distinctly from wrong guesses,
since ``I need more context'' on a personalized command is a different failure mode from
confidently emitting the wrong ecosystem. Neither touches the model or the memory mechanism ---
they are evaluation hygiene, and the earlier draft's retry-on-empty recommendation is now moot
given the reliable model produced no blanks.

\subsection{Corrective memory}

\paragraph{Impressive behavior.}
This is the paper's headline result, and the per-item data shows the mechanism cleanly ---
crucially, under the leakage-controlled protocol where the store holds only a \emph{paraphrase}
of each prior request (mean word-Jaccard 0.36 to the test query). With those reworded exemplars
retrieved by BM25 into the prompt, 23 of 24 project-specific items are correct, up from 9 correct
cold; 15 items flip from wrong-or-partial to correct. Table~\ref{tab:lift-examples} quotes
representative flips verbatim. Two sub-mechanisms are visible. The first is \textbf{ecosystem
correction}: a Rust or \code{uv}-native project answered with the commonest default
(\code{npm}, \code{eslint}, \code{shellcheck}) is corrected to the user's real toolchain. The
second is \textbf{exact-argument recall}: the model already had the right utility but the wrong
specifics, and memory supplies the precise config, path, or split the user expects reproduced
(\code{p02} ``run the training job'' gains \code{--config configs/base.yaml}; \code{p14} ``run
inference on the test set'' becomes \code{python infer.py --split test}).

\begin{table*}[t]
\centering
\footnotesize
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Request} & \textbf{Cold (no memory)} & \textbf{+\,Paraphrase memory} \\
\midrule
run the unit tests             & \texttt{npm test}                & \texttt{cargo test} \\
lint the code                  & \texttt{eslint .}                & \texttt{ruff check .} \\
build the release binary       & \texttt{make release}            & \texttt{cargo build --release} \\
install project dependencies   & \texttt{npm install}             & \texttt{uv sync} \\
check code formatting          & \texttt{shellcheck *.sh}         & \texttt{black --check .} \\
clean the build artifacts      & \texttt{rm -rf build/ dist/ \dots} & \texttt{cargo clean} \\
run the training job           & \texttt{python train.py}         & \texttt{python train.py --config configs/base.yaml} \\
deploy the model               & \textit{(asks for more context)} & \texttt{./scripts/deploy.sh --prod} \\
\bottomrule
\end{tabular}
\caption{Representative wrong/partial$\rightarrow$correct flips from the leakage-controlled
experiment (\code{lift.jsonl}). The store held only a \emph{paraphrase} of each prior request
(mean word-Jaccard 0.36) and the model was tested on the original, so a warm win requires
generalization across the paraphrase gap, not lookup. The top block is \emph{ecosystem
correction} (a Rust/\code{uv} project answered with \code{npm}/\code{eslint}/\code{shellcheck}
defaults); the last two rows are \emph{exact-argument recall} and \emph{abstention rescue}. All
warm outputs match gold.}
\label{tab:lift-examples}
\end{table*}

\paragraph{Unexpected errors.}
The paraphrase column has a single error out of 24, and it is the honest edge of the method:
\code{p19} ``run the benchmark suite'' is answered \code{cargo bench} where the user's real
command is \code{python -m benchmarks.run} --- the model resolved the ecosystem to Rust and the
retriever did not surface the paraphrased Python exemplar strongly enough to override it. This is
a retrieval-plus-prior miss, not a lookup failure, and it is exactly the residual one expects when
memory must generalize rather than copy. The \emph{shape} of the cold column is also informative:
of the 15 non-correct items, 9 are outright wrong and 6 partial, so memory is repairing a mix of
fundamentally-wrong ecosystem guesses and last-mile precision misses, not merely polishing
near-correct commands.

\paragraph{Phenomena.}
The most telling datum is the gap between the two warm conditions. Verbatim memory (the exact
prior request in the store) scores 100\% (24/24) --- the trivial lookup ceiling --- while
paraphrase memory scores 95\% (23/24). That the paraphrase condition falls only one item short of
verbatim lookup, despite a mean word-Jaccard of just 0.36 between the stored exemplar and the
query, is the clean measure of what BM25 buys here: the retriever bridges a substantial
rewording gap on 23 of 24 items, so the personalization win is a genuine generalization result
rather than an artifact of a planted answer. The one-item gap is the honest price of that
generalization.

\paragraph{Recommendations.}
The residual is a \emph{retrieval} miss, not a generation miss: when the paraphrase drifts far
enough and a strong ecosystem prior competes, lexical BM25 does not surface the stored pair
decisively. The cheapest next lever, given that stored commands now carry
\code{ts}/\code{session}/\code{cwd}, is recency-and-project weighting --- when several exemplars
tie lexically, prefer the user's \emph{recent} and \emph{same-directory} commands. Semantic
retrieval (embeddings) is the rung after that, warranted only once the domain broadens enough that
the paraphrase gap dominates, per \S6.

\paragraph{A note on token cost.}
Memory's effect on token cost is \emph{model-dependent}, an insight that cuts against a claim an
earlier draft of this work made. On non-reasoning claude-haiku-4-5, retrieved exemplars add a
small input cost (mean 95$\rightarrow$161 tokens/call, $+66$). But on a \emph{reasoning} model
(Qwen3-32B, from the free-tier runs of \S4.4) the same few-shot grounding suppresses the verbose
chain-of-thought the model otherwise emits when unsure, so net cost \emph{falls}
(385$\rightarrow$217 per call, $-44\%$). Memory as a token \emph{reduction} is thus a
reasoning-model phenomenon, not a universal property; we report the honest per-model effect
rather than headline the favorable case.

\subsection{Axis B: self-knowledge}

Docs memory lifts overall accuracy from 42\% to 75\% (35$\rightarrow$62 of 82), but the two
question types behave very differently, and the split is the story. Retrieval turns a plausible
guess into a grounded fact:

\begin{footnotesize}
\begin{verbatim}
k02 "How do I save the current file?"
 none: "Use the Save action."
       [would run: Save]              (acts, no key)
 docs: "Press C-x C-s to save
        the current buffer."          (correct)
r16 "How do I change the accent color?"
 none: "Mars doesn't have built-in
        theme color settings ..."     (wrong)
 docs: set theme_accent to an [r,g,b]
       array in tuning.json           (correct)
\end{verbatim}
\end{footnotesize}

\noindent The reconfigure slice is where memory is decisive: 16\% to \textbf{93\%}
(5$\rightarrow$28 of 30). Without docs the model invents plausible config locations ---
\code{r16} ``change the accent color'' is answered ``Mars doesn't have built-in theme color
settings,'' \code{r04} ``use a different model'' points at a \code{~/.mars/config.json} that does
not exist --- whereas the retrieved knob descriptions carry the exact name, file, and default, so
the agent answers \code{theme\_accent} / \code{MARS\_LLM\_MODEL} in \code{tuning.json} correctly.
Two failure families that motivated the corpus fix --- \textbf{acts-instead-of-explains}
(answering a how-to with a bare \code{[would run: \dots]} directive, \code{k02}, \code{k09}) and
\textbf{hallucinated identifiers} (a right-idea knob under the wrong filename) --- are largely
eliminated on this slice by the actionable, path-carrying knob descriptions and an
explain-don't-act instruction on the docs path.

Knowledge questions gain far less --- 57\% to 65\% --- for an instructive reason: they turn on an
\emph{exact keybinding}, and even with docs retrieved the model sometimes hallucinates a plausible
chord it prefers over the one the docs state (\code{k05} ``detach my session'' $\rightarrow$
\code{C-x C-c}, gold \code{C-x C-d}; \code{k06} ``search within a buffer'' $\rightarrow$
\code{C-f}, gold \code{C-s}; \code{k18} ``jump to top/bottom'' $\rightarrow$ \code{C-x Home}, gold
\code{M-<}). A knob \emph{name} is a token the retriever can hand over intact; a chord is a
specific fact the model will still overwrite with a more common default, which is why doc memory
helps reconfiguration far more than keybinding recall. A residual limit is judge slack: a few
correct-but-differently-phrased answers are marked wrong, so the reported knowledge number is a
mild lower bound even under the stronger claude-sonnet-5 oracle.

\end{document}
